### 2017-02-12 manuscript.rmd
modified tree parsing so nearly all are now time callibrated. Moved functions to separate file.

	#' Adds some random noise from a normal distribution to calibration
	#' times
	#' 
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @param sd_fraction this is multiplied by each value to set sd of that value 
	#' @return A dataframe as above, but with noise
	#' @export
	add_noise_calibration_times = function( calibration_times, sd_fraction ) {
		age = calibration_times$age
		new_age = NA

		# Keep trying until dates are coherent with tree structure
		# This is a ladder tree, so that means checking that they 
		# are still in ascending order
		while( NA %in% new_age ) {
			x = sapply( 
				age,
				function( i ){
					rnorm( 1, mean = i, sd = ( sd_fraction * i ) )
				}
			)

			if( ( ! is.unsorted( x ) ) & all( x > 0 ) ){
				new_age = x
			}

		}

		calibration_times$age = new_age
		return( calibration_times )
	}

	#' Get a boolean vector corresponding to all tips and internal nodes of 
	#' a tree, with value TRUE for tips and FALSE for internal nodes
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object
	#' @return A boolean vector
	#' @export
	is.tip.nhx = function( nhx ) {
		is.tip = rep( FALSE, nrow( nhx@data ) )
		is.tip[ 1:length( nhx@phylo$tip.label ) ] = TRUE
		is.tip
	}

	#' Fix ensembl clade names.
	#' Homininae is pan + homo + gorilla
	#' Hominini is pan + homo
	#' ENSEMBL Compara trees lislabel Hominini as Homininae
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object
	#' @return A phylogenetic tree as a treeio::treedata object
	#' @export
	fix_hominini = function( nhx ) {
		# Get the numbers for all nodes labeled as homininae
		homininae_nodes = nhx@data$node[ nhx@data$label == "Homininae" ]

		# Check which have no gorilla as a descendent
		no_gorilla = sapply(
			homininae_nodes,
			function( x ){
				tips = hutan::tip_descendants( nhx@phylo, x )
				tip_species = nhx@data$species[ tips ]
				return( ! "gorilla_gorilla" %in% tip_species )
			}
		)

		if( any( no_gorilla ) ){
			hominini_nodes = homininae_nodes[ no_gorilla ]
			nhx@data$label[ hominini_nodes ] = "Hominini"
		}

		return( nhx )
	}

	#' Record the distance from a tip for each node in a tree in 
	#' the data slot
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata
	#' @return A treeio::treedata object, with an additional node_age column
	#' in the data slot, or the original object if it was not of class treedata
	#' @export
	store_node_age = function( nhx ) {

		if ( class( nhx ) != "treedata" ) {
			return( nhx )
		}

		node_age = hutan::distance_from_tip( nhx@phylo )

		# make sure the dataframe is ordered by consecutive nodes
		stopifnot( all( nhx@data$node == 1:length( nhx@data$node ) ) )

		nhx@data$node_age = node_age

		return( nhx )
	}

	#' Get the number of speciation events in an nhx tree
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
	#' column @data$D that has value N for speciation events
	#' @return An integer indicating the number of speciation events
	#' @export
	get_n_speciation = function( nhx ){
		sum( nhx@data$D[ ! is.tip.nhx( nhx ) ] == "N", na.rm = TRUE )
	}

	#' Drop tips without expression data from a tree. 
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
	#' column @data$Tau containing expression data
	#' @param min_genes_with_expression The minimum number of tips with 
	#' expression data for a tree to be retained
	#' @return A treeio::treedata object, or NA if less than 
	#' min_genes_with_expression
	#' @export
	drop_empty_tips = function( nhx, min_genes_with_expression ) {

		# Identify the tips without expression data
		to_drop = which( is.na( nhx@data[ 1:length( nhx@phylo$tip.label ), ]$Tau ) )

		# Return the pruned tree if it meets sampling criteria, otherwise return NA
		remaining = length( nhx@phylo$tip.label ) - length( to_drop )
		if( remaining >= min_genes_with_expression ){
			pruned = treeio::drop.tip( nhx, to_drop )
			return( pruned )
		}
		else{
			return( NA )
		}
	}

	#' Adjust branch lengths of a phylogenetic tree to make it ultrametric and 
	#' to time calibrate the speciation nodes to fixed values
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with speciation 
	#' nodes annotated with a value of "N" in @data$D
	#' @param calibration_times A dataframe with two columns: age is the age 
	#' of a clade, and clade is the name of the clade
	#' @param ... Any additional arguments to pass to ape::chronos()
	#' @return A treeio::treedata object if successfully calibrated
	#' @export
	calibrate_tree = function ( nhx, calibration_times, ... ) {

		# Create calibration matrix for speciation nodes
		calibration = 
			nhx@data[ !is.tip.nhx( nhx ), ] %>%
			filter( D == "N" ) %>%
			left_join( calibration_times, c( "label" = "clade" ) ) %>%
			mutate( age.min = age ) %>%
			mutate( age.max = age ) %>% 
			select( node, age.min, age.max ) %>%
			mutate( soft.bounds = NA )

		tree = try( 
			ape::chronos( nhx@phylo, calibration=calibration, ... ) 
		)

		if( "phylo" %in% class( tree ) ){
			class( tree ) = "phylo"
			nhx@phylo = tree
			return( nhx )
		}
		else{
			return( NA )
		}
	}

	#' Calculate phylogenetic independent contrasts (PIC) with the tree and 
	#' character data in a treeio::treedata object
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object, with a 
	#' column @data$Tau containing expression data
	#' @return A treeio::treedata object, with new @data columns pic and var_exp
	#' @export
	pic.nhx = function( nhx ) {

		tau = nhx@data$Tau[ is.tip.nhx( nhx ) ]

		if( any( is.na(tau) ) ){
			stop( "A trait value is NA." )
		}

		# Calculate the contrasts
		p = pic( tau, nhx@phylo, var.contrasts=TRUE )

		# Add the results back to the @data slot, padding the rows that correspond to 
		# tips with NA
		nhx@data$pic = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,1 ] )
		nhx@data$var_exp = c( rep( NA, length( nhx@phylo$tip.label ) ), p[ ,2 ] )

		return( nhx )
	}

	#' Add phylogenetic independent contrasts (PIC) with the tree and 
	#' character data in a treeio::treedata object
	#' 
	#' @param gene_trees_calibrated A list of time calibrated phylogenetic trees 
	#' as treeio::treedata objects 
	#' @return A list of time phylogenetic trees with pics 
	#' as treeio::treedata objects
	#' @export
	add_pics_to_trees = function( gene_trees_calibrated ) {

		# Calculate independent contrasts for Tau on each tree, storing the results 
		# back into the @data slot of the tree objects
		gene_trees_pic = mclapply( gene_trees_calibrated, pic.nhx, mc.cores=cores )

		return( gene_trees_pic )
	}

	#' Add phylogenetic independent contrasts (PIC) with the tree and 
	#' character data in a treeio::treedata object
	#' 
	#' @param gene_trees_pic A list of phylogenetic trees with pics
	#' as treeio::treedata objects 
	#' @return A tibble with combined data, including phylogenetic independent 
	#' contrasts, for internal nodes of all trees
	#' @export
	summarize_contrasts = function( gene_trees_pic ) {
		# Combine the @data slots across all trees. These have labels, node annotations,
		# expression data, contrast results, etc...
		# Also take the absolute value of the contrast, since we only consider magnitude.
		nodes_all = 
			lapply( 
				gene_trees_pic, 
				function( nhx ){
					tags = nhx@data
					tags$gene = digest( nhx ) # Creates a hash that is unique to each gene tree
					tags %<>% select( -( B ) ) # B has inconsistent types, remove it
					return( tags )
				}
			) %>%
			bind_rows() %>%
			mutate( pic = abs( pic ) )

		# Create a subset that corresponds to nodes with contrasts. This excludes tips.
		nodes_contrast = 
			nodes_all %>%
			filter( ! is.na( pic ) ) %>%
			filter( ! is.na( D ) )

		return( nodes_contrast )
	}

	#' Calibrate a list of gene trees given a set of speciation node dates
	#' 
	#' @param gene_trees_pruned A list of phylogenetic trees as treeio::treedata objects 
	#' with associated character data
	#' @param calibration_times A dataframe with two columns: age.min is the age 
	#' of a clade, and clade is the name of the clade
	#' @param ... Any additional arguments to pass to ape::chronos()
	#' @return A list of calibrated phylogenetic trees as treeio::treedata objects. May be 
	#' shorter than gene_trees_pruned if some calibrations fail.
	#' @export
	calibrate_trees = function( gene_trees_pruned, calibration_times, ... ) {
		# Make the trees ultrametric and calibrate the speciation nodes to 
		# specified times
		gene_trees_calibrated = 
			mclapply( 
				gene_trees_pruned,
				calibrate_tree, 
				calibration_times=calibration_times,
				...,
				mc.cores=cores
			)

		# Remove trees that could not be successfully calibrated
		gene_trees_calibrated = gene_trees_calibrated[ ! is.na( gene_trees_calibrated ) ]

		# Parse the calibrated node ages from the internal @phylo object and store them with the 
		# corresponding rows in the @data object
		gene_trees_calibrated = 
			mclapply( 
				gene_trees_calibrated, 
				store_node_age, 
				mc.cores=cores 
			)

		return( gene_trees_calibrated )
	}

	#' Wilcoxon test of the ortholog conjecture, specifically that contrasts
	#' associated with duplication nodes are greater than those for 
	#' speciation nodes. A significant result indicates rejection of the null
	#' hypothesis that they are not greater.
	#' 
	#' @param nodes_contrast A tibble of independent contrasts
	#' @return Wilcoxon test p value
	#' @export
	wilcox_oc = function( nodes_contrast ) {

		p = wilcox.test( 
			nodes_contrast %>% filter( D=="Y" ) %>% .$pic, 
			nodes_contrast %>% filter( D=="N" ) %>% .$pic, 
			alternative="greater" 
		)$p.value

		return( p )
	}

	#' Estimate Tau evolution parameters from a tree::treedata tree
	#' and add them back into the tree object.
	#' 
	#' @param nhx A phylogenetic tree and associated Tau values 
	#' as a treeio::treedata object
	#' @param ... Additional arguments to pass to geiger::fitContinuous()
	#' @return A phylogenetic trees as a treeio::treedata object, 
	#' with estimated brownian motion model parameters stored 
	#' in @phylo$model
	#' @export
	add_model_parameters = function( nhx, ... ) {
		phy = nhx@phylo
		tau_original = nhx@data$Tau [ 1:length( phy$tip.label ) ]
		names( tau_original ) = phy$tip.label
		brownian_model = fitContinuous( phy, tau_original, ... )

		nhx@phylo$model$opt$z0 = brownian_model$opt$z0
		nhx@phylo$model$opt$sigsq = brownian_model$opt$sigsq

		return( nhx )
	}

	#' Simulate Tau on a tree::treedata. Existing observed values are
	#' fit to a brownian model to estimate parameters. These parameters
	#' are then used to replace the original values with similated 
	#' values.
	#' 
	#' @param nhx A phylogenetic tree and associated Tau values 
	#' as a treeio::treedata object
	#' @param dup_adjust A multiplier for adjusting the branch lengths 
	#' following duplication to effectively change the rate following duplication
	#' @return A phylogenetic trees and simulated Tau values as a 
	#' treeio::treedata object
	#' @export
	sim_tau = function( nhx, dup_adjust=1 ) {

		phy = nhx@phylo

		# Get parameter estimates
		brownian_model = nhx@phylo$model

		# Adjust length of branches that descend from duplication events. 
		# This is how we implement heterogeneous rates after duplication 
		# rather than speciation. Extending the branch lengths is equivalent 
		# to using a higher rate along them.
		dup_nodes = nhx@data$node[ nhx@data$D == "Y" ]
		dup_edges = phy$edge[,1] %in% dup_nodes
		phy$edge.length[ dup_edges ] = phy$edge.length[ dup_edges ] * dup_adjust

		# Simulate trait given the tree and parameter estimates
		x = fastBM( 
				phy, 
				a=brownian_model$opt$z0, 
				sig2=brownian_model$opt$sigsq, 
				bounds=c(-Inf,Inf) 
			) %>% abs()

		# Adjust the results so they fall between 0 and 1
		x[ x > 1 ] = 1
		x[ is.nan( x ) ] = 1

		names( x ) = NULL
		nhx@data$Tau = c( x, rep( NA, nrow( nhx@data ) - length (x ) ) )
		return( nhx )
	}

	#' Simulate a set of pics from a gamma distribution
	#' 
	#' @param scale Gamma scale parameter
	#' @param n Number of observations per event type	
	#' @return A dataframe
	#' @export
	sim_expectation = function( scale = 1, n = 10000 ){
		expectation = 
			bind_rows(
				data.frame( 
					pic = rgamma( n, 1, scale = scale ),
					Event = "Duplication",
					stringsAsFactors = FALSE
				),
				data.frame( 
					pic = rgamma( n, 1, scale = 1 ),
					Event = "Speciation",
					stringsAsFactors = FALSE
				)
			)

		event_levels = c( "Speciation", "Duplication" )
		expectation$Event = 
			factor( expectation$Event, levels=event_levels )

		return( expectation )
	}

	#' Summarize pairwise comparisons between tips of a tree
	#' 
	#' @param nhx A phylogenetic tree as a treeio::treedata object 
	#' 	
	#' @return A dataframe
	#' @export
	get_pairwise_summary = function ( nhx ) {

		# Create a tibble with one row for each pairwise combination of tip nodes
		ntips = length( nhx@phylo$tip.label )
		D = 1:ntips %>% combn( 2 ) %>% t() %>% as_tibble()
		names( D ) = c( "tip_a", "tip_b" )

		# Add a column with the gene tree hash 
		D %<>% mutate( gene_tree = digest( nhx ) )

		# Add the tip names
		D %<>% mutate ( name_a = nhx@data$G[ D$tip_a ] )
		D %<>% mutate ( name_b = nhx@data$G[ D$tip_b ] )

		# Add the species
		D %<>% mutate ( species_a = nhx@data$species[ D$tip_a ] )
		D %<>% mutate ( species_b = nhx@data$species[ D$tip_b ] )

		# Add the values at the tips
		D %<>% mutate ( tau_a = nhx@data$Tau[ D$tip_a ] )
		D %<>% mutate ( tau_b = nhx@data$Tau[ D$tip_b ] )
		D %<>% mutate ( difference = abs( tau_a - tau_b ) )

		# For each row, get the most recent common ancestor of the pairwise tip combination
		D$mrca = 
			apply(
				D, 
				1, 
				function( y ) ape::getMRCA( 
					nhx@phylo, 
					c( as.integer( y['tip_a'] ), as.integer( y['tip_b'] ) )
				)
			) %>%
			as.integer()

		# Merge data of this mrca to the node
		D %<>% left_join(
			nhx@data %>%
				select( D, Event, node, node_age, label, pic ),
			by = c( "mrca" = "node")

		)

		# Calculate the distance between the tips, which for an ultrametric tree 
		# is twice the age of their most recent common ancestor
		D %<>% mutate( distance = 2 * node_age )

		return( D )

	}

@@

	# Parse each string to a treeio::treedata object, adding a label column 
	# for node names
	gene_trees = mclapply(
		tree_lines,
		function( x ){
			tree_tc = textConnection( x )
			tree = treeio::read.nhx( tree_tc )
			close( tree_tc )
			tree@data$label = c( tree@phylo$tip.label, tree@phylo$node.label )
			return( tree )
		}, mc.cores=cores
	)

@@

			tree@data$Event = NA
			tree@data$Event[ tree@data$D == "N" ] = "Speciation"
			tree@data$Event[ tree@data$D == "Y" ] = "Duplication"
			tree@data$Event = factor( tree@data$Event, levels=c( "Speciation", "Duplication" ) )

>>>>

	source( "functions.R" )

@@

	# Parse each string to a treeio::treedata object, adding a label column 
	# for node names
	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )

@@
